---
title: "DESeq2 Tutorial - KEY"
author: Nicole Black
date: 10/10/20
---

```{r setup}
    knitr::opts_knit$set(root.dir = normalizePath("/Users/nicoleblack/Desktop/QBIO/qbio490_nicole/analysis_data")) 
```

```{r}
# install "DESeq2" using BiocManager::install()
BiocManager::install("DESeq2")

# load DESEq2, SummarizedExperiment, and TCGABiolinks
library(DESeq2)
library(SummarizedExperiment)
library(TCGAbiolinks)
```

```{r}
# you can read these in from .csv's or from your .RData file
rna_se <- READ IN
rna_counts <- READ IN
rna_genes <- READ IN
rna_clinical <- READ IN

load("/Users/nicoleblack/Desktop/QBIO/qbio490_nicole/analysis_data/rnaclinical_rnagenes_rnacounts.RData")
```

###### (1) Setting up and preprocessing our data

Firstly, we must convert the age_category column to a factor. Recall from the intro to R notebook that a factor is a categorical data type that stores values as well as different levels (a list of unique values). For example, within a column called 'grade' we might have the levels 'Freshman', 'Sophomore', 'Junior', 'Senior'.

Use factor() to convert rna_clinical$age_category into a factor.

```{r}
?factor

# write code here
rna_clinical$age_category <- factor(rna_clinical$age_category)
```

After factor conversion, check the levels by printing out the following:

```{r}
head(rna_clinical$age_category)
```

Note that in addition to the first few values, we now also see "Levels: old young". This indicates that rna_clinical$age_category was successfully converted into a factor.

In our analysis we can also control for variables that may skew our comparison. Let's control for stage at diagnosis (using ajcc_pathologic_stage) and one other variable of your choosing.

List your chosen variable here: paper_BRCA_Subtype_PAM50

Just like we did to age_category, convert these columns into the factor data type. Then call head and record the levels.

```{r}
# write code here
rna_clinical$ajcc_pathologic_stage <- factor(rna_clinical$ajcc_pathologic_stage)
rna_clinical$paper_BRCA_Subtype_PAM50 <- factor(rna_clinical$paper_BRCA_Subtype_PAM50)

# check levels with head
head(rna_clinical$ajcc_pathologic_stage) # Stages I through X
head(rna_clinical$paper_BRCA_Subtype_PAM50) # Basal, Her2, LumA, LumB, Normal
```

For DESeq2 to run, none of our variables can contain NAs. Just to see if there are NAs in age_category, ajcc_pathologic_stage, and your chosen variable by calling sum() on is.na() on the column.

```{r}
# write code here
sum(is.na(rna_clinical$age_category)) # 0
sum(is.na(rna_clinical$ajcc_pathologic_stage)) # 11
sum(is.na(rna_clinical$paper_BRCA_Subtype_PAM50)) # 19
```

If you get a non-zero number (aka there is at least one NA value in the column), create a boolean mask and get rid of any patients with an NA value. Make sure you delete these patients from both the rna_clinical data frame and the rna_counts data frame.

```{r}
na_mask <- !is.na(rna_clinical$ajcc_pathologic_stage) & !is.na(rna_clinical$paper_BRCA_Subtype_PAM50)
  
rna_clinical <- rna_clinical[na_mask, ]

rna_counts <- rna_counts[ , na_mask]
```

Now, let's move on to processing the rna_counts df. Here, we should remove all genes where the total number of counts (across all patients) is less than 10.

```{r}
# use rowSums() to create a list with the total number of counts of each gene
row_sums <- rowSums(rna_counts)

# create a boolean mask where genes with < 10 total counts are FALSE, and genes with >= 10 total counts are TRUE
low_counts_mask <- ifelse(row_sums < 10, F, T)

# rewrite the rna_counts df, subsetting for only genes with >= 10 total counts
rna_counts <- rna_counts[low_counts_mask, ]
```

Now we're ready for the analysis!

###### (2) Running DESeq2

To break down what analysis actually takes place within the DESeq2 pipeline:

1. The DESeqDataSetFromMatrix() function processes your transcript count data (rna_counts), the phenotypic data (rna_clinical), and the categories you want to compare on (the design argument) to create a DESeqDataSet object (dds). In a design formula the factor that follows "~" is called the group and it indicates the variable you want to split the data on (aka what clinical variables we are comparing in this analysis). The factor(s) that follow "+" are called the conditions and they indicates the variable(s) you want to control for.

2. The DESeq() function runs the comparisons and statistical analysis. This will likely take the bulk of the run time. Don't worry if this takes quite a while to run, just let it take its time.

3. The resultsNames() function will tell you what comparisons have been run. It’s helpful for when you have more than two levels in your experiment (in our case, there are just two, "young" and "old").

4. The results() function will make the comparisons readable. The contrasts argument calculates the fold change in "young" vs. "old", with the second level in the list being the denominator in the fold change (ie the baseline for the expression calculations).

```{r}
?DESeqDataSetFromMatrix
dds = DESeqDataSetFromMatrix(countData = rna_counts,
                             colData = rna_clinical,
                             design = ~age_category +ajcc_pathologic_stage +paper_BRCA_Subtype_PAM50)

?DESeq
dds_obj = DESeq(dds) # note: this will likely take a long time (ie 45 minutes to 2 hours)

?resultsNames
resultsNames(dds_obj)  # see what comparisons got run

# get the young vs. old comparison
?results
results = results(dds_obj, format = "DataFrame", contrast = c("age_category", "Young", "Old"))
```

###### (3) Analyzing and saving the data

First, explore the results df using head(), str(), and by double clicking on it in the Environment panel.

```{r}
# write code here
head(results)
str(results)
View(results)
```

We're not interested in all of these columns, so use data.frame() to rewrite results with the following columns in this order: gene name (from rna_genes), gene id (from results rownames), log2foldchange, pvalue, padj, -log10(padj). Rename the columns using colnames(). Note: to get -log10, simply write "-log10(fill in here)".

```{r}
# write code here
gene_mask <- ifelse(rna_genes$gene_id %in% rownames(results), T, F)
results <- data.frame(rna_genes$gene_name[gene_mask], rownames(results), results$log2FoldChange, results$pvalue, results$padj, -log10(results$padj))

# HINT: to get the gene names, you will need to create a boolean mask that is T when rna_genes$gene_id %in% results@rownames, and F if otherwise, then use this mask to subset rna_genes$gene_names for only the names that are in the results df

colnames(results) <- c("gene_name", "gene_id", "log2FC", "pvalue", "padj_value", "-log10padj_value")
```

It's oftentimes most interesting to look at genes that are highly over/under expressed at a significance level p < 0.05. 

First, select rows (genes) that have a padj value < 0.05.

```{r}
sig_results <- ifelse(is.na(results$padj_value), F, ifelse(results$padj_value < 0.05, T, F))
```

Now, sort the data frame by log2foldchange, descending. To do this, use the order() function. The order() function returns a vector of the sorted indexes of your chosen data, from smallest to largest. For example:

```{r}
my_df = data.frame(x = c('index 1', 'index 2', 'index 3', 'index 4', 'index 5'),
                   y = c(2,4,3,5,1))

order_indexes = order(my_df$y) # order by column "y"

# we would expect order_indexes to be c(5, 1, 3, 2, 4) since the value 2 has the index 5, the value 2 has the index 1, the value 3 has the index 3, etc.

order_indexes

my_df = my_df[order_indexes, ] # this rewrites the df based on the sorted rows
my_df # check your df to see if it is sorted by the y values
```

Use the logic above as well as the sig_results mask to sort the results df by log2foldchange descending. Note that this can also be done in one line by bypassing the use of order_indexes as an intermediate variable.

```{r}
up_reg_results <- results[sig_results, ]
order_indexes = order(up_reg_results$log2FC, decreasing = TRUE)
up_reg_results <- up_reg_results[order_indexes, ]
```

Select only the genes in up_reg_results where the log2foldchange is > 1 (ie genes that are approximately equally expressed between your populations).

```{r}
# write code here
up_reg_results <- up_reg_results[up_reg_results$log2FC > 1,]
```

Repeat the same steps to create a data frame called down_reg_results (this time select for all genes where log2foldchange is < -1). Note that you can set the decreasing argument to true in the order() function.

```{r}
# write code here
down_reg_results <- results[sig_results, ]
order_indexes = order(down_reg_results$log2FC)
down_reg_results<- down_reg_results[order_indexes, ]
down_reg_results<- down_reg_results[down_reg_results$log2FC < -1,]
```

Call head() on both up_reg_results_sorted and down_reg_results.

```{r}
# write code here
head(up_reg_results)
head(down_reg_results)
```

Notice that the row names are the old indexes from before we sorted. Create new row names that are more descriptive. 

```{r}
# write code here

row.names(up_reg_results) <- up_reg_results$gene_name 
row.names(down_reg_results) <- down_reg_results$gene_name 
```

Call head() again, and choose one up regulated (overexpressed) gene and one down regulated (underexpressed) gene and fill out the following information:

Gene ensembl id:
Log2FoldChange:
p-adj value:
More highly expressed in young or old patients:
Google the gene. What is its full name, and what appears to be its function? Genecards generally has a pretty good summary:

Gene ensembl id:
Log2FoldChange:
p-adj value:
More highly expressed in young or old patients:
Google the gene. What is its full name, and what appears to be its function? Genecards generally has a pretty good summary:

Write the results table to the week6 folder using write.csv().

```{r}
# write code here
write.csv(results, "results.csv")
write.csv(up_reg_results, "up_reg_results.csv")
write.csv(down_reg_results, "down_reg_results.csv")
```

###### (4) Creating volcano plots

The most common way to visualize differential expression results is with a volcano plot (named because it looks like a volcano erupting). A volcano plot shows the relative expression of each gene along with the significance for each gene's result.

The x-axis is log2FoldChange (ie the normalized fold change results), and the y axis is the -log10(p adjusted value).

First, think conceptually about which points in the volcano plot are the most interesting. Fill out the relative expression (up-regulated/down-regulated in young patients) and significance (yes/no) for the following locations on the volcano plot:

top-right genes: up reg, sig
bottom-right genes: up reg, not sig
top-left genes: down reg, sig
bottom-left genes: down reg, not sig
top-middle genes: approx equal, sig
bottom-middle genes: approx equal, not sig

There are many general plotting functions you can use to create a volcano plot (ex: plot(), gglot(), etc.), but we are going to use a specialized function made particularly for volcano plots as it allows for more customization. Install and load "EnhancedVolcano".

```{r}
# install (using BiocManager) and load "EnhancedVolcano"
BiocManager::install("EnhancedVolcano")
library(EnhancedVolcano)
```

Read the package documentation section 3.1, and create a volcano plot of your own.
https://www.bioconductor.org/packages/release/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html#plot-the-most-basic-volcano-plot

Remember that you may need to include par(mar=c(1,1,1,1)) before your plotting function to change the margin sizes.

```{r}
# create plot here
EnhancedVolcano(results,
    lab = results$gene_name,
    x = 'log2FC',
    y = 'padj_value')
```

Skim through section 4 of the package documentation, and create an advanced volcano plot as you see fit. Customize at least one aspect of your plot (ie color, shape, labels, etc.). Save this plot to your week6 folder as a .jpg file. For more info on saving plots, see section 3 of R_Tutorial.Rmd

```{r}
# create plot here
jpeg("../volcano_plot.jpg")
EnhancedVolcano(results,
    lab = results$gene_name,
    x = 'log2FC',
    y = 'padj_value',
    selectLab = c('TP53','PIK3CA', 'BRCA1', 'BRCA2'))
dev.off()
```

